Normalization by Evaluation for Sized Dependent Types

### 2017-08-13 nbe/icfp17.tex
[ ICFP 2017 ] break overfull box
[ ICFP 2017 ] fixed most issues raised by publisher

%% Title information
\title[NbE for Sized Types]{Normalization by Evaluation for Sized Dependent Types}
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Andreas Abel}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-0420-4492}             %% \orcid is optional
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Gothenburg University}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{andreas.abel@gu.se}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Andreas Vezzosi}
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Chalmers University of Technology}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{vezzosi@chalmers.se}          %% \email is recommended

\author{Theo Winterhalter}
\affiliation{
  \institution{École normale supérieure Paris-Saclay}
  \country{France}
}
\email{theo.winterhalter@gmail.com}

@@

Higher-rank size quantification takes termination checking to the next
level; it is necessary for abstraction and generic programming.  For
instance, it allows us to write a generic tree traversal which applies
a user-given preprocessor on subtrees before recursively descending
into these trees, and a postprocessor after surfacing from the
descent.
The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i)\; (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris~\cite{brady:jfp13}, and Haskell~\cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check needs special consideration, especially
when the equality-check is type directed.

@@

\section{A Type System With Irrelevant Size Application}
\label{sec:syntax}

@@

\subsection{Weak head reduction}

@@

\subsection{Read back}
\label{sec:read}

@@

\subsection{Partial equivalence relations}
\label{sec:per}

@@

\subsection{PER model}
\label{sec:permodel}

@@

\subsection{Type shapes}
\label{sec:shape}

@@

\subsection{Computation with natural numbers}
\label{sec:comp}

@@

\SHORTVERSION{\clearpage}

>>>>

%% Title information
\title{Normalization by Evaluation for Sized Dependent Types}
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Andreas Abel}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-0420-4492}             %% \orcid is optional
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Gothenburg University, Sweden}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{andreas.abel@gu.se}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Andreas Vezzosi}
\affiliation{
  \department{Department of Computer Science and Engineering}
  \institution{Chalmers University of Technology, Sweden}
  \streetaddress{Rännvägen 6b}
  \city{Göteborg}
%  \state{State1}
  \postcode{41296}
  \country{Sweden}
}
\email{vezzosi@chalmers.se}          %% \email is recommended

\author{Theo Winterhalter}
\affiliation{
  \institution{École normale supérieure Paris-Saclay, France}
  \country{France}
}
\email{theo.winterhalter@gmail.com}

@@

Higher-rank size quantification takes termination checking to the next
level; it is necessary for abstraction and generic programming.  For
instance, it allows us to write a generic tree traversal which applies
a user-given preprocessor on subtrees before recursively descending
into these trees, and a postprocessor after surfacing from the
descent.
The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i)\; (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris~\cite{brady:jfp13}, and Haskell~\cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t =\\
  \phantom{\ttrav\,}\vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check needs special consideration, especially
when the equality-check is type directed.

@@

\section{A Type System with Irrelevant Size Application}
\label{sec:syntax}

@@

\subsection{Weak Head Reduction}

@@

\subsection{Read Back}
\label{sec:read}

@@

\subsection{Partial Equivalence Relations}
\label{sec:per}

@@

\subsection{PER Model}
\label{sec:permodel}

@@

\subsection{Type Shapes}
\label{sec:shape}

@@

\subsection{Computation with Natural Numbers}
\label{sec:comp}

@@

%%\SHORTVERSION{\clearpage}
